Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-04-05 tex/abstract.tex
intro and abstract

% \begin{abstract}
\todo{refine / update}
As machine learning models increase in complexity, the human ability of understanding and interpreting decisions made by those models has not been able to keep up.
The usefulness of white-box analysis techniques, exposing the internal state of models, are limited to relatively simple models and have trouble with complex models, such as deep neural networks.
Recently, black-box machine learning analysis techniques offer model independent insights into the decision making process of machine learning models.
For quickly and effectively gaining insights from those techniques, visual analytics emerges as a powerful set of tools.
We use visual analytics to explore both global and local means of explaining and understanding predictive models via black-box techniques.
We then propose a model diagnostic workflow that uses aggregated instance level explanations to overcome problems of fully global or local methods.
That is, by avoiding global aggregates, finer details of the decision making process are retained, while going beyond individual instances, analysts are not overwhelmed by the quantity of instances to inspect.
Finally, we show that the model diagnostic workflow can not only help improving the models themselves but offers insights about flaws in the \emph{input data}, thus helping with the task of feature engineering.

>>>>

% \begin{abstract}
As machine learning models increase in complexity, the human ability of understanding and interpreting decisions made by those models has not been able to keep up.
The usefulness of white-box analysis techniques, exposing the internal state of models, are limited to relatively simple models and have trouble with complex models, such as deep neural networks.
Recently, black-box machine learning analysis techniques offer model independent insights into the decision making process of machine learning models.
For quickly and effectively gaining insights from those techniques, visual analytics emerges as a powerful set of tools.
We use visual analytics to explore both global and local means of explaining and understanding predictive models via black-box techniques.
We then propose a model diagnostic workflow that uses aggregated instance-level explanations to overcome problems of fully global or local methods.
That is, by avoiding global aggregates, finer details of the decision making process are retained, while going beyond individual instances, analysts are not overwhelmed by the quantity of instances to inspect.
Finally, we show that the Model Diagnostic workflow can not only help improving the models themselves but offers insights about flaws in the \emph{input data}, thus helping with the task of feature engineering.

### 2018-04-05 tex/introduction/contribution.tex
intro and abstract

\section{Research Overview and Thesis Contributions}
The main goal of this thesis is to explore how model agnostic feature contribution methods can help to gain a holistic understanding of, as well as, insights about predictive models using visual analytics. In three stages, we will show the progression from initial observations, that led us to utilizing black-box analysis, to the eventual use of aggregated instance-level explanations to understand, trust, and verify predictive models. Additionally, we show that our developed visual analytics techniques can be helpful for feature engineering.

>>>>

\section{Research Overview and Thesis Contributions}
\label{sec:corecontributions}
The main goal of this thesis is to explore how model agnostic feature contribution methods can help to gain a holistic understanding of, as well as, insights about predictive models using visual analytics. In three stages, we will show the progression from initial observations, that led us to utilizing black-box analysis, to the eventual use of aggregated instance-level explanations to understand, trust, and verify predictive models. Additionally, we show that our developed visual analytics techniques can be helpful for feature engineering.

### 2018-04-05 tex/introduction/introduction.tex
intro and abstract

\todo{State core contributions}

\todo{outline of introduction}

>>>>

With the growing complexity of machine learning models and the need for algorithm agnostic approaches, black-box explanations offer insights into the external behavior of such models.
Visual analytics can be used to leverage the full potential of those explanations in a scalable way.

This thesis explores both global and local approaches for black-box explanations.
Furthermore, it shows the advantage of the compromise of both techniques: aggregating instance-level explanations.
This results in the Model Diagnostic workflow which not only helps improving machine learning models, but also offers insights about problems with the input data, assisting in the task of feature engineering.

Following, we will discuss essentials in both machine learning, starting in Section~\ref{sec:machinelearning}, and visual analytics, in Section~\ref{sec:visualanalytics}.
Then, in Section~\ref{sec:usecase}, we will briefly discuss the common use case, of machine learning for health care, within this thesis.
Afterwards, we will motivate the work of the thesis in Section~\ref{sec:motivation_intro} followed by an overview of the research and contributions of the thesis in Section~\ref{sec:corecontributions}.
In Section~\ref{sec:thesisoutline} we will then outline the remainder of this thesis.

### 2018-04-05 tex/introduction/machine_learning.tex
much more text
some additions
intro and abstract

\section{Machine Learning and Predictive Modeling}
Machine learning enables computers to perform tasks without manually codifying all instructions.
This automated programming makes machine learning a popular and widely used approach for many applications as it provides a time- and cost-effective alternative to manually created software solutions.
Machine learning relies on algorithms that learn and generalize from examples in order to perform accurately on new or unseen data.

@@

@@

@@

\input{figs/roc}

The \emph{receiver operating characteristic} (ROC) curve for binary classifiers plots the true positive rate against the false positive rate for all thresholds between $0$ and $1$.
The plot indicates the correctness of the model \wrt its confidence.
The ROC curve for a perfect model ($100\%$ accuracy) would follow the left vertical axis and the top horizontal axis.
The area under the ROC curve (\emph{AUC}) is a commonly used measure for model quality \cite{bradley19971145}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{tex/decisiontree}
    \caption[Example Decision Tree.]{A possible decision tree for the Diabetes prediction example\footnotemark. Note, that the tree has interval checks to model the relationship between ``weight'' in \si{kg} and ``height'' in \si{m}.}
    \label{figs:dctree}
\end{figure}
\footnotetext{The decision tree or its parameters are \emph{not} based on data and serve only illustratory purposes.}

@@

\par \noindent \textbf{Biased Data:}
Biased data might occur if the acquired data stems from a sub-population that insufficiently represents the entire population.
For example, if we train our Diabetes model from above only with obese people the resulting model will most likely have a greater tendency to predict Diabetes.

@@

\section{Visual Analytics}
Gaining meaningful insights about large data sets is hard.
While computing various statistics about the data is often helpful it might oversimplify or in some cases even mislead information about the data.
One example data set often used to demonstrate this is Anscombe's quartet~\cite{doi:10.1080/00031305.1973.10478966}.
The quartet consists of four different data sets that all have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
However, plotting the values of the data sets reveals that each one of those data sets obeys a different, unique characteristic (see Figure~\ref{figs:anscombe}).

\begin{figure*}[t]
\centering
\includegraphics[width=0.9\textwidth,valign=t]{tex/introduction/anscombe.png}
\caption[Anscombe's quartet.]{
Plots of the four data sets of Anscombe's quartet\footnotemark. All data sets have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
}
\label{figs:anscombe}
\end{figure*}
\footnotetext{Image source: Wikipedia}

@@

>>>>

\section{Machine Learning and Predictive Modeling}
\label{sec:machinelearning}
Machine learning enables computers to perform tasks without manually codifying all instructions.
This automated programming makes machine learning a popular and widely used approach for many applications as it provides a time- and cost-effective alternative to manually created software solutions.
Machine learning relies on algorithms that learn and generalize from examples in order to perform accurately on new or unseen data.

@@

\begin{figure}
\centering
\includegraphics[height=8em]{figs/cm}
\caption[Example of a confusion matrix.]{
Example of a confusion matrix. Each cell represents the number of instances under a certain condition. Rows show the ground truth while columns show the predicted outcomes from the machine learning model.
For example, there were 44 instances predicted ``high'' whose ground truth is actually ``low''.
The edge of the matrix shows the tallies of the rows and columns.
}
\label{figs:cm_intro}
\end{figure}

@@

All of those measures can be read from a confusion matrix, as shown in Figure~\ref{figs:cm_intro}

@@

\begin{figure}
\centering
\includegraphics[height=10em]{figs/roc}
\caption[Example of a ROC curve.]{
Example of a ROC curve. The plot indicates the correctness of the model with the false-positive rate as horizontal axis and the true-positive rate as vertical axis. The line traces those values for different prediction thresholds. The threshold minimizing incorrect predictions is highlighted by the colored axes.
}
\label{figs:roc_intro}
\end{figure}

The \emph{receiver operating characteristic} (ROC) curve for binary classifiers plots the true positive rate against the false positive rate for all thresholds between $0$ and $1$, as shown in Figure~\ref{figs:roc_intro}.
The plot indicates the correctness of the model \wrt its confidence.
The ROC curve for a perfect model ($100\%$ accuracy) would follow the left vertical axis and the top horizontal axis.
The area under the ROC curve (\emph{AUC}) is a commonly used measure for model quality \cite{bradley19971145}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{tex/decisiontree}
    \caption[Example Decision Tree.]{A possible decision tree for the Diabetes prediction example\footnotemark. Note, that the tree has interval checks to model the relationship between ``weight'' in \si{kg} and ``height'' in \si{m}.}
    \label{figs:dctree}
\end{figure}
\footnotetext{The tree and its parameters are \emph{not} based on any data and serve only illustratory purposes.}

@@

\par \noindent \textbf{Biased Data:}
Biased data might occur if the acquired data stems from a sub-population that insufficiently represents the entire population.
For example, if we train our Diabetes model from above only with obese people the resulting model will most likely have a greater tendency to predict Diabetes.
Note, that imputing missing values might also result in a biased data set.

@@

\section{Visual Analytics}
\label{sec:visualanalytics}
Gaining meaningful insights about large data sets is hard.
While computing various statistics about the data is often helpful it might oversimplify or in some cases even mislead information about the data.
One example data set often used to demonstrate this is Anscombe's quartet~\cite{doi:10.1080/00031305.1973.10478966}.
The quartet consists of four different data sets that all have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
However, plotting the values of the data sets reveals that each one of those data sets obeys a different, unique characteristic (see Figure~\ref{figs:anscombe}).

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\textwidth,valign=t]{tex/introduction/anscombe.png}
\caption[Anscombe's quartet.]{
Plots of the four data sets of Anscombe's quartet\footnotemark. All data sets have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
}
\label{figs:anscombe}
\end{figure}
\footnotetext{Image source: Wikipedia}

### 2018-04-05 tex/introduction/motivation.tex
intro and abstract

\section{Motivation}
Modern machine learning algorithms have reached a degree of potential that lets the resulting models model almost any data set.
But even though there are statistical measures to guarantee that the model is learning correctly without over-fitting on the training data, the model with the highest accuracy is not always the most usable model or even semantically correct.
The following three examples aim to showcase this:

>>>>

\section{Motivation}
\label{sec:motivation_intro}
Modern machine learning algorithms have reached a degree of potential that lets the resulting models model almost any data set.
But even though there are statistical measures to guarantee that the model is learning correctly without over-fitting on the training data, the model with the highest accuracy is not always the most usable model or even semantically correct.
The following three examples aim to showcase this:

### 2018-04-05 tex/introduction/usecase.tex
intro and abstract

\section{Use Case: Machine Learning for Health Care}
In this thesis we will mostly explore use cases from the medical domain\footnote{However, results are easily transferable to other domains as well.}.
It is quite interesting to see parallels between the decision making process of doctors and machine learning.

>>>>

\section{Use Case: Machine Learning for Health Care}
\label{sec:usecase}
In this thesis we will mostly explore use cases from the medical domain\footnote{However, results are easily transferable to other domains as well.}.
It is quite interesting to see parallels between the decision making process of doctors and machine learning.
